---
title: "rddj-template"
subtitle: "My analysis"
author: "My Name"
date: "October 2018"
output:
  html_document:
    code_folding: show
    df_print: kable
    toc: yes
    toc_depth: 4
    toc_float: 
      collapsed: false
      smooth_scroll: false
---


```{r, echo=FALSE}
# CONFIG
user_name <- "grssnbchr" # your Git username (only needed if
# you want to deploy to GH pages)
project_name <- "rddj-template" # adapt!
package_date <- "2018-09-01" # date of the CRAN snapshot that
# the checkpoint package uses
R_version <- "3.5.2" # R-Version to use
options(Ncpus = 4) # use 4 cores for parallelized installation of packages
if (R_version != paste0(version$major, ".", version$minor)){
  stop("ERROR: specified R version does not match currently used.")
}
```

## Notes

This report was generated on `r Sys.time()`. R version: `r paste0(version$major, ".", version$minor)` on `r version$platform`. For this report, CRAN packages as of `r package_date` were used.

...

### R-Script & data

The preprocessing and analysis of the data was conducted in the [R project for statistical computing](https://www.r-project.org/). The RMarkdown script used to generate this document and all the resulting data can be downloaded [under this link](http://`r user_name`.github.io/`r project_name`/rscript.zip). Through executing `main.Rmd`, the herein described process can be reproduced and this document can be generated. In the course of this, data from the folder `ìnput` will be processed and results will be written to `output`. 

### GitHub

The code for the herein described process can also be freely downloaded from [https://github.com/`r user_name`/`r project_name`](https://github.com/`r user_name`/`r project_name`). 


### License

...

### Data description of output files

#### `abc.csv` (Example)

| Attribute | Type | Description |
|-------|------|-----------------------------------------------------------------------------|
| a | Numeric | ... |
| b | Numeric | ... |
| c | Numeric |  ... |

### `xyz.csv` 

...

## Preparations

```{r, echo=FALSE}
detach_all_packages <- function() {
  basic_packages_blank <-  c("stats",
                             "graphics",
                             "grDevices",
                             "utils",
                             "datasets",
                             "methods",
                             "base")
  basic_packages <- paste("package:", basic_packages_blank, sep = "")

  package_list <- search()[
    ifelse(unlist(gregexpr("package:", search())) == 1, TRUE, FALSE)]

  package_list <- setdiff(package_list, basic_packages)

  if (length(package_list) > 0)  for (package in package_list) {
    detach(package, character.only = TRUE, unload = TRUE)
    print(paste("package ", package, " detached", sep = ""))
  }
}

detach_all_packages()

# this allows multiple persons to use the same RMarkdown
# without adjusting the working directory by themselves all the time
source("scripts/csf.R")
path_to_wd <- csf() # if this - for some reason - does not work,
# replace with a hardcoded path, like so: "~/projects/rddj-template/analysis/"
if ( is.null(path_to_wd) | !dir.exists(path_to_wd)) {
  print("WARNING: No working directory specified for current user")
} else {
  setwd(path_to_wd)
}

# suppress scientific notation
options(scipen = 999)

# unload global rstudioapi and knitr again to avoid conflicts with checkpoint
# this is only necessary if executed within RStudio
# outside of RStudio, namely in the knit.sh script, this causes RMarkdown
# rendering to fail, thus should not be executed there
if (Sys.getenv("RSTUDIO") == "1"){
  detach_all_packages()
}
```


### Define packages

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# from https://mran.revolutionanalytics.com/web/packages/checkpoint/vignettes/using-checkpoint-with-knitr.html
# if you don't need a package, remove it from here (commenting is probably not sufficient)
# tidyverse: see https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/
cat("
library(rstudioapi)
library(tidyverse) # ggplot2, dplyr, tidyr, readr, purrr, tibble
library(magrittr) # pipes
library(readxl) # excel
library(scales) # scales for ggplot2
library(jsonlite) # json
library(lintr) # code linting
library(sf) # spatial data handling
library(rmarkdown)",
file = "manifest.R")
```

### Install packages

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# if checkpoint is not yet installed, install it (for people using this
# system for the first time)
if (!require(checkpoint)) {
  if (!require(devtools)) {
    install.packages("devtools", repos = "http://cran.us.r-project.org")
    require(devtools)
  }
  devtools::install_github("RevolutionAnalytics/checkpoint",
                           ref = "v0.3.2", # could be adapted later,
                           # as of now (beginning of July 2017
                           # this is the current release on CRAN)
                           repos = "http://cran.us.r-project.org")
  require(checkpoint)
}
# nolint start
if (!dir.exists("~/.checkpoint")) {
  dir.create("~/.checkpoint")
}
# nolint end
# install packages for the specified CRAN snapshot date
checkpoint(snapshotDate = package_date,
           project = path_to_wd,
           verbose = T,
           scanForPackages = T,
           use.knitr = F,
           R.version = R_version)
rm(package_date)
```


### Load packages

```{r, echo=TRUE, message=FALSE, warning=FALSE}
source("manifest.R")
unlink("manifest.R")
sessionInfo()
```

### Load additional scripts

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# if you want to outsource logic to other script files, see README for 
# further information
knitr::read_chunk("scripts/my_script.R")
source("scripts/my_script.R")
my_function(5)
```

## IO test

```{r, echo=TRUE, message=FALSE, warning=FALSE}
x <- read.csv2("input/test.csv", sep = ",")
x
write.csv2(x, "output/abc.csv", row.names = F)
write.csv2(x, "output/xyz.csv", row.names = F)
# the following file would be ignored by Git if saved into the ignore folder
# see README for further information
# write.csv2(x, "output/ignore/fgh.csv", row.names = F)
```

## Some plot

```{r, echo=TRUE, message=FALSE, warning=FALSE}
set.seed(123)
x <- rnorm(1000)
y <- x * 2 + rnorm(1000)
df <- data.frame(x, y)
ggplot(df) +
  geom_point(aes(x = x, y = y)) +
  geom_smooth(aes(x = x, y = y)) +
  labs(title = "The correlation between x and y", 
       subtitle = "A ggplot2 approach", 
       caption = "source: a pseudo-number generator with seed 123") +
  theme_minimal()
``` 

## Linting

The code in this RMarkdown is listed with the [lintr package](https://github.com/jimhester/lintr), which is based on the  [tidyverse style guide](http://style.tidyverse.org/). 

```{r echo=TRUE, message=FALSE, warning=FALSE}
lintr::lint("main.Rmd")
# if you have additional scripts and want them to be linted too, add them here
# lintr::lint("scripts/my_script.R")
```
